---
title: "Getting started: honest intervals under nonlinear confounding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: honest intervals under nonlinear confounding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, fig.align = "center")
library(bllnn)
```

## The problem

You have a treatment or exposure whose effect you want to estimate, and a pile
of covariates that influence both it and the outcome. The relationship between
those covariates and the outcome is not linear, and you have no particular
theory about its shape.

$$y = x\beta + f(z) + \varepsilon$$

The target is $\beta$, with an interval you can defend. Everything about
$f(z)$ is a nuisance -- but a nuisance that will contaminate $\beta$ if you
model it badly, and contaminate it in a way that no amount of extra data
repairs.

Fitting a linear model and hoping is the usual failure. Let us watch it fail.

```{r simulate}
sim <- sim_partial_linear(n = 300, beta = c(treat = 1.5), p_z = 5,
                          confounding = 0.6, seed = 11)
str(sim$data, give.attr = FALSE)
```

The truth is $\beta = 1.5$. The covariates `z1`--`z5` drive the outcome through
a smooth nonlinear function, and they are also correlated with `treat` -- that
is what `confounding = 0.6` sets up.

```{r ols}
ols <- lm(y ~ treat + z1 + z2 + z3 + z4 + z5, data = sim$data)
round(coef(ols)[["treat"]], 4)
```

```{r ols-hidden, echo = FALSE}
b_ols <- coef(ols)[["treat"]]
ci_ols <- confint(ols)["treat", ]
```

Least squares returns `r round(b_ols, 3)` against a truth of 1.5, with a 95%
confidence interval of [`r round(ci_ols[1], 3)`, `r round(ci_ols[2], 3)`].
The interval is narrow, confident, and wrong: the curvature in $f(z)$ that the
linear terms cannot absorb ends up in the treatment coefficient, because
`treat` is the covariate best positioned to absorb it.

Note what does *not* fix this. The bias is not sampling noise, so a larger
sample does not shrink it; it converges to the wrong number.

## The fit

```{r fit}
fit <- bllnn(y ~ z1 + z2 + z3 + z4 + z5, data = sim$data,
             linear = ~ treat, folds = 3, n_iter = 800, burn = 200,
             seed = 11)
summary(fit)
```

```{r fit-hidden, echo = FALSE}
b_hat <- coef(fit)[["treat"]]
ci <- confint(fit)["treat", ]
```

The estimate is `r round(b_hat, 3)` with a 95% credible interval of
[`r round(ci[1], 3)`, `r round(ci[2], 3)`], which contains the truth. Note the
interval is *wider* than the least-squares one. That is the point rather than a
cost: the narrow interval was narrow because the model was certain about
something false.

The formula splits the two roles explicitly. The left side of `linear = ~ treat`
names the terms you want coefficients for; everything in the main formula is
handed to the network as nuisance. Nothing is inferred about which is which.

## What makes the interval honest

Two things, and they are not interchangeable with anything else.

**The features are frozen.** The network body is learned by cross-fitting --
trained on held-out folds, then fixed before any sampling happens. This is
sample splitting, not a performance shortcut. Once the features are fixed, the
conditional posterior of the last layer is exactly Gaussian and is drawn in
closed form, so the block is an exact Gibbs step rather than an approximation
embedded in one.

**The linear terms are residualised against the nuisance.** This is the part
that is easy to get backwards, and getting it backwards is worse than skipping
it.

The network is deliberately given the confounding channel: `bllnn_crossfit()`
trains an auxiliary body estimating $E[x \mid z]$ and appends it to the
features. But that creates a second problem -- the features can now represent
$E[x \mid z]$ exactly, while $y$ depends on it only through $x\beta$, so the
network and the coefficient compete for the same direction and $\beta$ is
attenuated. The fix is to hand the sampler $x - E[x \mid z]$ instead of $x$.
This is Robinson partialling-out, the identifying step behind double machine
learning, and `bllnn()` does it for you.

The direction matters. Projecting the *features* onto the orthogonal
complement of $x$ -- the superficially similar operation -- forces
$X'\Phi = 0$, which removes the network from the coefficient draw entirely and
reduces $\beta$ to ordinary least squares with the confounding still in it.
Measured over 25 simulated datasets, that variant covered the truth 8% of the
time against a nominal 95%.

## Checking the chain

This is MCMC, so the output is a chain and deserves to be looked at rather
than trusted.

```{r diagnostics, fig.alt = "Trace and posterior density for the treatment coefficient"}
plot(fit)
```

The trace should look like noise around a level, with no drift and no long
excursions. The effective sample size in the summary above is the number that
matters for interval accuracy; a few hundred is enough for a credible interval,
and if it is much smaller, raise `n_iter` before believing the interval.

## What this is really for

The wrapper above is a convenience. `bllnn` is a *component*: the reason it
exists is that you can put the draw inside a sampler of your own, alongside
whatever else your model needs. That is the subject of the companion vignette,
`vignette("custom-sampler", package = "bllnn")`, which also covers binary and
count outcomes.

## One caution

The credible interval for $\beta$ is the claim this package makes and defends.
Pointwise credible bands for $f(z)$ itself are a different quantity and are
**not** honest in the same way: they describe uncertainty about the best fit
within the span of the frozen features, and say nothing about how far that
span sits from the truth. On a smooth test problem, nominal 95% bands for
$f(z)$ covered it 58% of the time. This is a property of the frozen-feature
design rather than of any particular posterior, and it does not affect
$\beta$, which rests on residualisation rather than on $f$ being represented
exactly.
