---
title: "Using bllnn inside your own Gibbs sampler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using bllnn inside your own Gibbs sampler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bllnn)
```

The `bllnn()` wrapper fits one model: a partial-linear one, with a Gaussian
outcome. This vignette is about the thing underneath it, which fits nothing at
all and is more useful for that reason.

## The contract

A `bllnn_sampler` is one conditional draw, packaged so you can call it from a
loop you wrote. It mirrors `dbarts`:

| you call | it means |
|---|---|
| `set_response(mod, r)` | here is the residual you are responsible for |
| `set_sigma(mod, s)` | here is the current noise **standard deviation** |
| `gibbs_step(mod)` | give me one draw of `f`, as a vector of length `n` |

The object has reference semantics. `set_response()` and `set_sigma()` modify
it in place, so nothing is reassigned each sweep.

Two things it will never do. It will not estimate the noise scale -- that
belongs to you, and estimating it here would use the data twice and break the
joint chain. And it will not silently run an invalid kernel: `gibbs_step()`
errors unless `is_valid_kernel(mod)` is `TRUE`.

## A complete host sampler

The partial-linear model $y = x\beta + f(z) + \varepsilon$, written out by
hand. This is what `bllnn()` does internally, and there is nothing hidden in
it.

```{r setup-data}
sim <- sim_partial_linear(n = 400, beta = c(treat = 1.5), p_z = 5,
                          confounding = 0.6, seed = 3)
y <- sim$data$y
x <- as.matrix(sim$data[, "treat", drop = FALSE])
z <- as.matrix(sim$data[, paste0("z", 1:5)])
```

Learn the features once, by cross-fitting, and freeze them. Supplying
`linear = x` also trains an auxiliary body for $E[x \mid z]$, which
`partial_out()` then uses to residualise.

```{r build}
cf  <- bllnn_crossfit(z, y, linear = x, folds = 3, seed = 3)
mod <- bllnn_sampler(cf, tau2 = "auto")
xr  <- partial_out(cf)            # x - E[x | z]

is_valid_kernel(mod)
```

Now the loop. Three conditionals, of which `bllnn` supplies one.

```{r gibbs}
set.seed(3)
n_iter <- 600; burn <- 200
p <- ncol(xr)
XtX <- crossprod(xr)
prior <- diag(p) / 100
beta <- rep(0, p)
sigma <- sd(y)
keep <- matrix(NA_real_, n_iter, p)

for (s in seq_len(n_iter)) {
  ## f | beta, sigma  -- the block's job
  set_response(mod, y - as.vector(xr %*% beta))
  set_sigma(mod, sigma)
  f <- gibbs_step(mod)

  ## beta | f, sigma  -- yours
  r <- y - f
  V <- solve(XtX / sigma^2 + prior)
  m <- V %*% (crossprod(xr, r) / sigma^2)
  beta <- as.vector(m + t(chol(V)) %*% rnorm(p))

  ## sigma^2 | beta, f  -- also yours
  resid <- r - as.vector(xr %*% beta)
  sigma <- sqrt(1 / rgamma(1, 2 + length(y) / 2, 1 + sum(resid^2) / 2))

  keep[s, ] <- beta
}

post <- keep[-seq_len(burn), , drop = FALSE]
round(c(mean = mean(post), quantile(post, c(0.025, 0.975))), 3)
```

```{r gibbs-hidden, echo = FALSE}
pm <- mean(post); pq <- quantile(post, c(0.025, 0.975))
```

The posterior mean is `r round(pm, 3)` with a 95% interval of
[`r round(pq[1], 3)`, `r round(pq[2], 3)`], against a truth of 1.5.

Note the residual handed over each sweep is `y - x_resid %*% beta`, using the
*residualised* design. That is not cosmetic: `Phi` contains $E[x \mid z]$ by
construction, so if the raw `x` were used the network and $\beta$ would
compete for the same direction and $\beta$ would be attenuated toward
$\beta(1 - \rho^2)$.

## Binary outcomes

For a logistic outcome there is no residual scale -- you cannot subtract your
contribution from a 0/1 response. So the host passes its contribution as an
*offset* instead, and the block conditions on it. Set
`posterior = "polyagamma"`, which is exact rather than approximate:
conditional on Polya-Gamma latent variables the logistic likelihood is
Gaussian in the weights.

```{r binary-data}
set.seed(7)
n <- 600
zb <- matrix(runif(n * 3, -1, 1), n, 3)
fb <- 1.5 * sin(pi * zb[, 1]) - zb[, 2]^2
xb <- rbinom(n, 1, 0.5)
yb <- rbinom(n, 1, plogis(0.8 * xb + fb))
```

```{r binary-fit}
body <- bllnn_warmup(zb, yb - 0.5, width = 16, epochs = 150, seed = 7)
modb <- bllnn_sampler(body, tau2 = "auto", posterior = "polyagamma",
                      data = zb)
set_response(modb, yb)          # the outcome itself, not a residual
```

The host's own step needs the latent variables, which `gibbs_step()` leaves on
the object as `modb$omega`. Given `omega` the outcome is conditionally
Gaussian with observation precisions `omega`, so the coefficient draw is an
ordinary weighted normal update.

```{r binary-gibbs}
nb <- 300; b <- 0; keepb <- numeric(nb)

for (s in seq_len(nb)) {
  set_offset(modb, b * xb)      # the host's contribution to the linear predictor
  eta <- gibbs_step(modb)
  om <- modb$omega

  V <- 1 / (sum(om * xb^2) + 1 / 25)
  m <- V * sum(xb * ((yb - 0.5) - om * eta))
  b <- rnorm(1, m, sqrt(V))
  keepb[s] <- b
}

kb <- keepb[-seq_len(100)]
round(c(mean = mean(kb), quantile(kb, c(0.025, 0.975))), 3)
```

```{r binary-hidden, echo = FALSE}
bm <- mean(kb); bq <- quantile(kb, c(0.025, 0.975))
```

`r round(bm, 3)` with a 95% interval of [`r round(bq[1], 3)`,
`r round(bq[2], 3)`], against a truth of 0.8.

## Counts

Negative-binomial counts use the same augmentation with a different exponent.
Set `dispersion = r`, a positive integer, and pass non-negative counts to
`set_response()`. The dispersion is **not** estimated: the augmentation needs
`PG(y + r, psi)`, which is exact only for whole `r`, so profile over a few
values rather than expecting the sampler to learn it.

Poisson counts are a different matter. The Polya-Gamma identity needs the
$(1 + e^\psi)^{-b}$ form, which the Poisson likelihood does not have, so the
exact augmentation cannot reach it. For that case there is
`laplace_moments()`, which is approximate and **not** a valid kernel --
`gibbs_step()` refuses `posterior = "laplace"` unless you pass
`force = TRUE`. Use it one-shot, not in a loop.

```{r laplace}
set.seed(5)
Phi <- matrix(rnorm(300 * 4), 300, 4)
yp <- rpois(300, exp(as.vector(Phi %*% c(0.4, -0.3, 0.2, 0))))

post_l <- laplace_moments(Phi, yp, tau2 = 10, family = "poisson")
round(post_l$mean, 3)
```

How good that approximation is has been measured rather than assumed, against
a long Polya-Gamma run as ground truth. Both its errors shrink with `n`: the
gap between the mode and the posterior mean falls from 0.45 posterior standard
deviations at `n` = 60 to 0.14 at `n` = 1000, and interval width from 0.95 of
the exact width to 1.00 by `n` = 500. Where it is wrong it is overconfident.
See `inst/validation/laplace_checks.R`.

## The guard

Not every combination is a legitimate transition kernel, and the ones that are
not will produce plausible output while targeting the wrong distribution. That
is the worst failure mode available here, so it is refused by default.

```{r guard, error = TRUE}
bad <- bllnn_sampler(Phi, tau2 = 1, posterior = "bootstrap")
is_valid_kernel(bad)

set_response(bad, rnorm(300))
set_sigma(bad, 1)
gibbs_step(bad)
```

`valid_kernels()` lists every combination and why each is or is not valid.

```{r table}
valid_kernels()[, c("posterior", "features", "valid")]
```
