---
title: "A worked example: the LaLonde job training data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A worked example: the LaLonde job training data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_matchit <- requireNamespace("MatchIt", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_matchit)
library(bllnn)
```

```{r no-matchit, eval = !has_matchit, echo = FALSE, results = "asis"}
cat("**This vignette needs the MatchIt package**, which supplies the data and",
    "is not installed. Install it and rebuild to see the analysis.\n")
```

The simulations elsewhere in this package have a known truth, which is what
makes them useful and also what makes them easy. This vignette runs `bllnn` on
a real dataset where the truth is contested, the assumptions are visibly
violated, and the answer is less comfortable.

It is also, deliberately, the example that found a defect. That part is kept
in.

## The data

The LaLonde data are the standard difficult case in causal inference. The
treated units are men who entered the National Supported Work Demonstration, a
mid-1970s job training program. The controls are **not** the experimental
controls -- they are drawn from the Population Survey of Income Dynamics, a
general population survey. The outcome is 1978 earnings.

The data ship with `MatchIt`, and are referenced here rather than copied.

```{r data}
data("lalonde", package = "MatchIt")
str(lalonde)
```

Two features make this hard, and both are worth seeing before any model runs.

```{r hard}
## 1. The groups barely overlap. Pre-treatment earnings differ enormously.
round(tapply(lalonde$re74 == 0, lalonde$treat, mean), 3)

## 2. The outcome has a large point mass at exactly zero.
round(mean(lalonde$re78 == 0), 3)
```

Seventy-one percent of the treated had zero earnings in 1974, against
twenty-six percent of the controls. These are not comparable populations, and
no adjustment method conjures comparability that the data do not contain.

Twenty-three percent of the outcomes are exactly zero. Earnings are
semicontinuous and right-skewed, so the Gaussian likelihood `bllnn()` assumes
is wrong here. It is used anyway, because that is what practitioners do with
this dataset and because the alternative would obscure the point of the
example -- but it is a real caveat, not a technicality.

## What the raw comparison says

```{r naive}
round(diff(tapply(lalonde$re78, lalonde$treat, mean)), 0)
```

The trained group earned about \$635 *less*. Taken at face value the program
looks harmful, which is the whole reason this dataset is famous: the
comparison is dominated by the fact that the two groups were different people
to begin with.

```{r linear}
lin <- lm(re78 ~ treat + age + educ + race + married + nodegree + re74 + re75,
          data = lalonde)
round(c(coef(lin)[["treat"]], confint(lin)["treat", ]), 0)
```

Linear adjustment moves the estimate to about \$1,548 and the interval only
just excludes zero. For reference, Dehejia and Wahba (1999) report an
experimental benchmark of roughly \$1,794 from the randomised comparison --
a number that comes from data not in this file, so treat it as a landmark
rather than as ground truth for what follows.

## The defect this example found

Running `bllnn()` with its defaults gives something absurd, and it is
instructive rather than embarrassing.

```{r default-prior}
fit_default <- bllnn(re78 ~ age + educ + race + married + nodegree + re74 + re75,
                     data = lalonde, linear = ~ treat, folds = 3,
                     n_iter = 1500, burn = 500, seed = 1)
round(c(coef(fit_default)[["treat"]], confint(fit_default)["treat", ]), 1)
```

An effect of a few dollars, with an interval a few tens of dollars wide, for an
outcome whose standard deviation is over \$7,000. That is not an estimate; it
is the prior.

The cause is `prior_beta`, whose default is 100 -- a prior standard deviation
of \$10 on the treatment coefficient. In the package's simulations the outcome
has a standard deviation of about 5, so a prior of `N(0, 100)` is diffuse and
harmless. On earnings in dollars it asserts, before seeing anything, that the
effect is almost certainly between -\$20 and \$20. The data contribute a
precision of roughly `sum(xr^2)/sigma^2`, about `1e-6` here, against the
prior's `0.01`; the prior outweighs the likelihood by a factor of thousands
and the posterior is simply the prior.

**`prior_beta` is not scale-free, and you must set it to match your outcome.**
A defensible default is the variance of the response, which says the effect is
unlikely to exceed one outcome standard deviation by much.

```{r scaled-prior}
fit <- bllnn(re78 ~ age + educ + race + married + nodegree + re74 + re75,
             data = lalonde, linear = ~ treat, folds = 3,
             n_iter = 1500, burn = 500,
             prior_beta = var(lalonde$re78), seed = 1)
summary(fit)
```

```{r hidden, echo = FALSE}
bh <- coef(fit)[["treat"]]; cb <- confint(fit)["treat", ]
```

## What to conclude

```{r compare, echo = FALSE}
data.frame(
  method = c("raw difference", "linear adjustment", "bllnn"),
  estimate = round(c(diff(tapply(lalonde$re78, lalonde$treat, mean)),
                     coef(lin)[["treat"]], bh)),
  lower = round(c(NA, confint(lin)["treat", 1], cb[1])),
  upper = round(c(NA, confint(lin)["treat", 2], cb[2]))
)
```

`bllnn` puts the effect at about \$`r round(bh)` with a 95% credible interval
of [\$`r round(cb[1])`, \$`r round(cb[2])`]. Three things are worth saying
about that plainly.

**The sign is corrected.** Flexible adjustment moves the estimate from
-\$635 to something clearly positive, in the neighbourhood of both the linear
estimate and the experimental landmark.

**The interval includes zero, and the linear one does not.** This is not
`bllnn` performing worse. The linear model buys its narrow interval by
assuming the covariates enter linearly; when that assumption is relaxed, the
evidence for a nonzero effect turns out to be weaker than the linear fit
suggests. An honest method producing a wider interval on genuinely
hard-to-adjust data is the intended behaviour, not a shortfall.

**It does not settle the question.** With overlap this poor, no method
recovers the experimental answer reliably, and this example should not be read
as a claim that `bllnn` does. What it demonstrates is that the interval
responds to the difficulty of the problem rather than hiding it.

## Caveats, stated once and plainly

- The Gaussian likelihood is wrong for this outcome: 23% of values are exactly
  zero and the rest are right-skewed. A hurdle or two-part model would be
  better specified, and this package does not currently offer one.
- The experimental benchmark comes from a comparison not present in these
  data.
- Pointwise credible bands for `f(z)` are not honest in the way the
  coefficient interval is; see the limitations section of the README.
- `prior_beta` must be set on the scale of your outcome. This vignette exists
  partly to make that impossible to miss.

## References

Lalonde, R. (1986). Evaluating the econometric evaluations of training programs
with experimental data. *American Economic Review* 76: 604-620.

Dehejia, R. H. and Wahba, S. (1999). Causal effects in nonexperimental studies:
re-evaluating the evaluation of training programs. *Journal of the American
Statistical Association* 94: 1053-1062.
