---
title: "A worked example: the LaLonde job training data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A worked example: the LaLonde job training data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_matchit <- requireNamespace("MatchIt", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_matchit)
library(bllnn)
```

```{r no-matchit, eval = !has_matchit, echo = FALSE, results = "asis"}
cat("**This vignette needs the MatchIt package**, which supplies the data and",
    "is not installed. Install it and rebuild to see the analysis.\n")
```

The simulations elsewhere in this package have a known truth, which is what
makes them useful and also what makes them easy. This vignette runs `bllnn` on
a real dataset where the truth is contested, the assumptions are visibly
violated, and the answer is less comfortable.

It is also the example that found two defects in this package, both of them
invisible to every simulation here because simulated responses are conveniently
scaled. Those are fixed, and the account of them is kept in rather than tidied
away, because the reasoning transfers to whatever you fit next.

## The data

The LaLonde data are the standard difficult case in causal inference. The
treated units are men who entered the National Supported Work Demonstration, a
mid-1970s job training program. The controls are **not** the experimental
controls -- they are drawn from the Population Survey of Income Dynamics, a
general population survey. The outcome is 1978 earnings.

The data ship with `MatchIt`, and are referenced here rather than copied.

```{r data}
data("lalonde", package = "MatchIt")
str(lalonde)
```

Two features make this hard, and both are worth seeing before any model runs.

```{r hard}
## 1. The groups barely overlap. Pre-treatment earnings differ enormously.
round(tapply(lalonde$re74 == 0, lalonde$treat, mean), 3)

## 2. The outcome has a large point mass at exactly zero.
round(mean(lalonde$re78 == 0), 3)
```

Seventy-one percent of the treated had zero earnings in 1974, against
twenty-six percent of the controls. These are not comparable populations, and
no adjustment method conjures comparability that the data do not contain.

Twenty-three percent of the outcomes are exactly zero. Earnings are
semicontinuous and right-skewed, so the Gaussian likelihood `bllnn()` assumes
is wrong here. It is used anyway, because that is what practitioners do with
this dataset and because the alternative would obscure the point of the
example -- but it is a real caveat, not a technicality.

## What the raw comparison says

```{r naive}
round(diff(tapply(lalonde$re78, lalonde$treat, mean)), 0)
```

The trained group earned about \$635 *less*. Taken at face value the program
looks harmful, which is the whole reason this dataset is famous: the
comparison is dominated by the fact that the two groups were different people
to begin with.

```{r linear}
lin <- lm(re78 ~ treat + age + educ + race + married + nodegree + re74 + re75,
          data = lalonde)
round(c(coef(lin)[["treat"]], confint(lin)["treat", ]), 0)
```

Linear adjustment moves the estimate to about \$1,548 and the interval only
just excludes zero. For reference, Dehejia and Wahba (1999) report an
experimental benchmark of roughly \$1,794 from the randomised comparison --
a number that comes from data not in this file, so treat it as a landmark
rather than as ground truth for what follows.

## Why the prior is scaled to your data

```{r fit}
fit <- bllnn(re78 ~ age + educ + race + married + nodegree + re74 + re75,
             data = lalonde, linear = ~ treat, folds = 3,
             n_iter = 1500, burn = 500, seed = 1)
summary(fit)
```

```{r hidden, echo = FALSE}
bh <- coef(fit)[["treat"]]; cb <- confint(fit)["treat", ]
```

That works on the defaults, and it is worth saying why it did not always, because
the reason generalises past this dataset.

`prior_beta` is the prior variance on the linear coefficients, and it cannot be
a fixed number. A coefficient multiplies `treat` to produce dollars, so any
statement about its plausible size is a statement in units of
`sd(y) / sd(treat)`. Until 2026-09-07 the default was a flat 100 -- a prior
standard deviation of \$10 -- which is diffuse when the response has a standard
deviation of about 5, as everything in this package's simulations does, and
absurd when the response is earnings. On this data the likelihood contributed a
precision near `1e-6` against the prior's `0.01`, so the prior won by three
orders of magnitude and the reported posterior was the prior: an effect of one
dollar, with an interval of [-\$19, \$22], on an outcome whose standard
deviation exceeds \$7,000. Nothing warned.

The default is now `"auto"`, which sets `(2.5 * sd(y) / sd(x_j))^2` per
coefficient -- the convention `rstanarm` uses, wide enough that the likelihood
dominates. You can still pass a number, and if you do, it is on you to put it on
the scale of your outcome.

The same example exposed a second scale dependence, in the network rather than
the prior. `bllnn_warmup()` centred the response but did not scale it, and Adam
moves the weights by about `learn_rate` per step whatever the gradient is -- so a
fixed epoch budget buys a fixed distance in weight space while the distance
needed grows with the response. The body fitted here was never arriving. The
response is now standardised, and both are checked by tests that assert
invariance under a change of units rather than a threshold at one scale, because
a threshold at one scale is exactly what failed to catch either of them.

## What to conclude

```{r compare, echo = FALSE}
data.frame(
  method = c("raw difference", "linear adjustment", "bllnn"),
  estimate = round(c(diff(tapply(lalonde$re78, lalonde$treat, mean)),
                     coef(lin)[["treat"]], bh)),
  lower = round(c(NA, confint(lin)["treat", 1], cb[1])),
  upper = round(c(NA, confint(lin)["treat", 2], cb[2]))
)
```

`bllnn` puts the effect at about \$`r round(bh)` with a 95% credible interval
of [\$`r round(cb[1])`, \$`r round(cb[2])`]. Three things are worth saying
about that plainly.

**The sign is corrected.** Flexible adjustment moves the estimate from
-\$635 to something clearly positive, in the neighbourhood of both the linear
estimate and the experimental landmark.

**The interval includes zero, and the linear one does not.** This is not
`bllnn` performing worse. The linear model buys its narrow interval by
assuming the covariates enter linearly; when that assumption is relaxed, the
evidence for a nonzero effect turns out to be weaker than the linear fit
suggests. An honest method producing a wider interval on genuinely
hard-to-adjust data is the intended behaviour, not a shortfall.

**It does not settle the question.** With overlap this poor, no method
recovers the experimental answer reliably, and this example should not be read
as a claim that `bllnn` does. What it demonstrates is that the interval
responds to the difficulty of the problem rather than hiding it.

## Caveats, stated once and plainly

- The Gaussian likelihood is wrong for this outcome: 23% of values are exactly
  zero and the rest are right-skewed. A hurdle or two-part model would be
  better specified, and this package does not currently offer one.
- The experimental benchmark comes from a comparison not present in these
  data.
- Pointwise credible bands for `f(z)` are not honest in the way the
  coefficient interval is; see the limitations section of the README.
- The prior and the network now both scale themselves to your response, but
  nothing scales your *judgement* of what a plausible effect is. If you set
  `prior_beta` by hand, set it in the units of your outcome.

## References

Lalonde, R. (1986). Evaluating the econometric evaluations of training programs
with experimental data. *American Economic Review* 76: 604-620.

Dehejia, R. H. and Wahba, S. (1999). Causal effects in nonexperimental studies:
re-evaluating the evaluation of training programs. *Journal of the American
Statistical Association* 94: 1053-1062.
